Population PME Project Proposal

### 2023-07-12 paper/proposal.tex
Updates from Overleaf
update proposal

\title{Population PME Project Propsal}
\author{Robert Zielinski}
\date{July 12, 2023}

@@

The modeling approach I am considering right now is influenced by the Latent Trajectory Model in \cite{schulamIntegrativeAnalysisUsing2016}. Essentially, rather than using a single spline function to represent the embedding function, I consider using several spline functions, each representing a different level of the model in question. For example, if we were interested in using the ADNI data to model the hippocampus at one time point, we may consider representing the embedding as a sum of three spline functions. One spline would represent the population level, another would represent the subgroup level (for instance, whether someone belongs to the cognitively healthy group or the group with AD), and the final spline would represent individual-level variability. The number of splines used, and their structure, would ultimately be flexible depending on the situation being modeled. However, I believe the general structure could be flexible enough to adapt to a number of distinct scenarios.

@@

\begin{itemize}
  \item Data Reduction
  \item Initialization: Run PME on data from one observation, then use the estimated manifold to find parameterizations for all other observations.
  \item Fitting: Iterate between estimating spline coefficients in alignment with the model structure determined for the given problem and updating the parameterization for each observation
  \item Tuning: For each spline function, select the most appropriate smoothing value
\end{itemize}

The specific approach for data reduction, as well as how the concentration of points across individuals will be calculated, is currently unclear. Additionally, the estimation mechanism for the spline coefficients is also unclear. In \cite{schulamIntegrativeAnalysisUsing2016}, the EM algorithm is used to estimate the parameters, so this may be a viable option in this case as well.

@@

We have discussed using the MNIST dataset as a possible example dataset, and I think this would be a good case for it. We could consider trying to find an aggregate estimate of a manifold for the figure "0", where differences in the shape of the character from the group-wide estimate would be handled by the random-effects type of structure described above.

>>>>

\title{Population PME Project Proposal}
\author{Robert Zielinski}
\date{July 12, 2023}

@@

The modeling approach I am considering right now is influenced by the Latent Trajectory Model in \cite{schulamIntegrativeAnalysisUsing2016}. Essentially, rather than using a single spline function to represent the embedding function, I consider using several spline functions, each representing a different level of the model in question. For example, if we were interested in using the ADNI data to model the hippocampus at one time point, we may consider representing the embedding as a sum of three spline functions. One spline would represent the population level, another would represent the subgroup level (for instance, whether someone belongs to the cognitively healthy group or the group with AD), with the spline modeling any differences between the population and each subgroup, and the final spline would represent individual-level variability from the subgroup they belong to. The number of splines used, and their structure, would ultimately be flexible depending on the situation being modeled. However, I believe the general structure could be flexible enough to adapt to a number of distinct scenarios. Because the hierarchical structure is over a large number of spline coefficients that each interact with each other, the additive approach will be helpful for keeping the model flexible, and having coefficients clearly assigned to different groups of interest will improve interpretability.

@@

\begin{itemize}
  \item Initialization: Run PME on data from one observation, then use the estimated manifold to find parameterizations for all other observations.
  \item Data Reduction: Run HDMDE on each observation individually.
  \item Fitting: Iterate between estimating spline coefficients in alignment with the model structure determined for the given problem and updating the parameterization for each observation
  \item Tuning: For each spline function, select the most appropriate smoothing value
\end{itemize}

The estimation mechanism for the spline coefficients is currently unclear. In \cite{schulamIntegrativeAnalysisUsing2016}, the EM algorithm is used to estimate the parameters, so this may be a viable option in this case as well.

The situation where we account for longitudinal change becomes more complicated.

@@

We have discussed using the MNIST dataset as a possible example dataset, and I think this would be a good case for it. We could consider trying to find an aggregate estimate of a manifold for the figures "0", "1", "3", "5", and "7", where differences in the shape of the character from the group-wide estimate would be handled by the random-effects type of structure described above. The model structure in this case would have two levels, one for the population level and one for the individual level variations. If we wanted to make the situation more complex, we could also divide images of the figures "2", "4", and possibly "9" into the various methods people use to write them. These different writing methods could then represent subgroups, which would be represented by another level in the model structure.
